Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-05-24 submission/Sze_Obesity_PeerJ_2016.Rmd
fixing spacing in text

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05). When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`). When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 4), nsmall=4)`) between non-obese and obese individuals. Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

@@

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.  We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al. study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al. Study).  For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al. study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al. study). The total number of trees and node size was standardized to 500 and 20 respectively for each data set. Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).  One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study. The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8), Bacteroides (5/8), and Shannon diversity metric (4/8 studies).

@@

We next moved on to the pooled meta-analysis. For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A]. The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`). In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`)[Figure 3B].  The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).  The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].

@@

Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A]. Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`.  However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`, Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`using this approach, between obese and non-obese individuals [Figure 5B].

>>>>

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05). When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`). When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals. Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

@@

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.  We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al. study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al. Study).  For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al. study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al. study). The total number of trees and node size was standardized to 500 and 20 respectively for each data set. Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).  One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study. The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8), Bacteroides (5/8), and Shannon diversity metric (4/8 studies).

@@

We next moved on to the pooled meta-analysis. For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A]. The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`). In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].  The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).  The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].

@@

Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A]. Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`).  However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) using this approach, between obese and non-obese individuals [Figure 5B].
